---
title: "DATA 698 - Final Project"
author: "Calvin Wong"
date: "5/12/2021"
output:
  pdf_document: default
  html_document: default
---

```{r message=FALSE, warning=FALSE, include=FALSE}

library(DT)
library(dplyr)
library(tidyverse)
library(sp)
library(xlsx)
library(stringr)
library(lubridate)
library(data.table)
library(tidyr)
library(factoextra)
library(cluster)
library(nomclust)

#Load dataset
peo <- read.csv("https://raw.githubusercontent.com/cwong79/DATA698/master/peo2.csv", na.strings="NA")

#Assigning Longitude and Latitude points
coordinates <- SpatialPointsDataFrame(peo[,c('Longitude', 'Latitude')] , peo)
```

## Introduction

In this final project, I will focus on a co-employment arrangement called a Professional Employer Organization (PEO) in the State of Florida, in which a business organization outsources its HR, employee benefits, and payroll functions to a PEO provider that specializes in these functions. My goal is to understand why business organizations select PEO providers and what drivers impacts that selection.

## Research Question

In this project I will attempt to answer the following research questions:

* What important criterias cause a business organization to select a specific PEO provider?  
* What competitive business advantage can PEO providers derive from this understanding to improve their service offerings?


###Project Background

As an employee of ADP (Automated Data Processing), a comprehensive global provider of human capital management (HCM) solutions, my research is to develop a competitive sales research to provide my organization a sales advantage within the PEO provider market. Fortune Business Insights says that the global human capital management (HCM) market size was USD 16.24 billion in 2019 and is projected to reach USD 32.68 billion by 2027. Like most HCM providers, ADP's diverse products seek to unite human resources, payroll distribution, talent management, time and labor management, tax and benefits administration, business outsourcing services, work analytics and compliance expertise into a single software platform.

One of ADP’s HR outsourcing products is called TotalSource PEO. PEO’s (professional employer organizations) partner with companies to provide comprehensive HR outsourcing to help manage a company’s human resources, employee benefits, regulatory compliance and payroll outsourcing. TotalSource is the largest certified professional employer organization in the United States, and now co-employs more than half a million worksite employees. ADP TotalSource has continued to grow organically over the past 19 years. In fact, if ADP TotalSource were an independent employer, its more than 500,000 worksite employees would rank it second among private sector employers in the United States.

A PEO works through a co-employment arrangement, which means the PEO contractually shares certain employer responsibilities and risk with the company. The contract between a client and the Professional Employer Organization (PEO) establishes the co-employment relationship. The PEO assumes certain aspects of employer responsibilities. Your employees become co-employed by both you and the PEO, thereby giving you access to the PEO’s built-in HR, compliance, benefits, a workers' compensation insurance program, payroll and tax infrastructure.

####NAICS 

The North American Industry Classification System (NAICS) is the standard used by Federal statistical agencies in classifying business establishments for the purpose of collecting, analyzing, and publishing statistical data related to the U.S. business economy.

NAICS was developed under the auspices of the Office of Management and Budget (OMB), and adopted in 1997 to replace the Standard Industrial Classification (SIC) system. It was developed jointly by the U.S. Economic Classification Policy Committee (ECPC), Statistics Canada, and Mexico's Instituto Nacional de Estadistica y Geografia, to allow for a high level of comparability in business statistics among the North American countries.

Please refer to https://www.census.gov/naics/?58967?yearbck=2017 for the latest codes and associated business classification. NAICS classification plays a big role in PEO providers growth strategies. This is because PEO's are formulated to absorb deficiencies business organizations incur as a result of operating within their industries. For example, PEO provider, Southeast Leasing traditionally services blue-collar industries which perform non-technical service-based work such as General Construction services and Hospitality-based industries. This is because of Southeast Leasing’s capabilities around managing Workers’ Compensation risk through its carrier Lion Insurance. Lion Insurance is focused on higher injury risk industries and offers specific programs to offset those risks of its insurers. Another example, PEO provider, Insperity excels at servicing technical white-collar industries which perform technical work within Technology and Finance. Their offering assist in securing high medical coverages at lower costs, because in those industries, medical offerings are used as an incentive to attract and retain that talent pool. The PEO market at a higher extent comprises of PEO providers looking to provide niche offerings to specific NAICS industries as a means of business competitive advantage. 

PEO tiers is an unknown PEO industry designation created to provide basic categorization of PEOs. The tiers fundamentally divide PEOs based on costs, with Tier 1 PEOs typically being higher consumer cost options. It is important to note that these designations do not just correlate with cost but also service levels and other factors to differentiate between tiers.

####Survey Rating

The National Association of Professional Employer Organizations (NAPEO) is The Voice of the PEO Industry. NAPEO's mission is to create and cultivate a legislative and regulatory climate at the federal level and in all 50 states that recognizes the key role PEOs play in supporting small and mid-size businesses and positions the PEO industry for continued growth. In 2018, NAPEO sent out a survey to all PEO users to ascertain how the factors of Health Benefits, Workers’ Compensation, Business strategy, Cost savings and Human Resource Support played in a business decision to utilize a PEO. The survey place a ranking from one to five with five being the highest level of importance the factor played in the business decision to join the PEO. 

```{r message=FALSE, warning=FALSE, include=FALSE}

#Create NAICS class type based on first 2 numeric
peo$NAICSClassCode <- as.numeric(substr(peo$NAICS, start = 1, stop = 2))

type <- c("Agriculture, Forestry, Fishing and Hunting", 
          "Mining", 
          "Utilities", 
          "Construction", 
          "Manufacturing", 
          "Wholesale Trade", 
          "Retail Trade",
          "Transportation and Warehousing", 
          "Information", 
          "Finance and Insurance", 
          "Real Estate Rental and Leasing", 
          "Professional, Scientific, and Technical Services", 
          "Management of Companies and Enterprises", 
          "Administrative and Support and Waste Management and Remediation Services", 
          "Educational Services", 
          "Health Care and Social Assistance", 
          "Arts, Entertainment, and Recreation", 
          "Accommodation and Food Services", 
          "Other Services (except Public Administration)", 
          "Public Administration")
naicsdata <- data.frame(type)
peo$NAICS_TYPE <- cut(peo$NAICSClassCode, c(1, 11, 21, 22, 23, 34, 42, 46, 50, 51, 52, 53, 54, 55, 57, 61, 63, 71, 73, 82, Inf), type)

#Unknown NAICS class
peo$NAICS_TYPE <- as.character(peo$NAICS_TYPE)
peo$NAICS_TYPE[is.na(peo$NAICS_TYPE)] <- "Unknown"

#Convert to date format
peo$LOCATION.EFFECTIVE.DATE <- as.Date(peo$LOCATION.EFFECTIVE.DATE, "%m/%d/%y")
peo$month <- month(peo$LOCATION.EFFECTIVE.DATE)
peo$RenewalMonth <- format(peo$LOCATION.EFFECTIVE.DATE, "%m")
choices_month <- seq.Date(from = as.Date("1/1/19"), by = "month", length.out = 36)

#Creating a group called Tier 1
tier1 <- c("ADP TOTAL SOURCE INC", "A 1 HR A DIVISION OF OASIS OUTSOURCING INC", "COADVANTAGE CORP", "INSPERITY INC", "OASIS ACQUISITION INC", "OASIS ACQUISITION INC A PAYCHEX CO", "OASIS DHR LLC", "OASIS OUTSOURCING CONTRACT II INC", "OASIS OUTSOURCING INC", "PAYCHEX BUSINESS SOLUTIONS LLC", "PAYCHEX HR OUTSOURCING LLC", "TRINET GROUP INC", "TRINET HR II HOLDINGS INC", "TRINET HR IV LLC")

#Creating a group called Tier 2
tier2 <- c("ALLY HR LLC DBA MATRIXONESOURCE", "ALPHASTAFF GROUP INC", "CHOICE EMPLOYER SOLUTIONS INC", "CORNERSTONE CAPITAL GROUP INC", "DECISION HR", "FLORIDA RESOURCE MANAGEMENT LLC", "FRANKCRUM 2 INC", "IMPACT STAFF LEASING LLC", "JUSTWORKS EMPLOYMENT GROUP LLC", "KYMBERLY GROUP PAYROLL SOLUTIONS INC", "OCMI III INC DBA PEOPAYGO", "REGIS GROUP HOLDINGS INC", "SOUTH EAST PERSONNEL LEASING INC", "STAFFLINK OUTSOURCING INC", "THE S2 HR GROUP LLC", "TLR OF BONITA INC", "WORKFORCE BUSINESS SERVICES INC")

#Partitioning the groups
peo$Tier <- with(peo, ifelse(NAMED.INSURED %in% tier1, "1", 
                             ifelse(NAMED.INSURED %in% tier2, "2", "3")))
```

### Data Source

The dataset that I will use for this project is a publicly available dataset from the State of Florida Workers’ Compensation Department, which contains workers comp coverage data over 2018 and 2019.  
The dataset is sourced from the Florida workers comp website (https://www.myfloridacfo.com/division/wc/), and has been downloaded and saved as a raw CSV file at:

https://raw.githubusercontent.com/cwong79/DATA698/master/peo2.csv

This dataset includes 10,730 cases of 47 variables, with each case relating to a specific company's workers comp insurance policy.  The 25 variables include policy information (policy dates, insurance agent, carrier, etc.) as well as insured company information (name, location, number of employees, etc.). The survey ranking was merged into this database and contained within HealthRating, WCRating, Strategy, Cost and Support fields. 

For purpose of this project, our target variable will be derived from `Tier`.  

```{r}
glimpse(peo)
```

## Methodology

In this project I will use clustering analysis to determine what PEO user share similar characteristics. Cluster analysis, can be defined as an unsupervised machine learning technique that aims to find patterns while gathering data samples and group them into similar records using predefined distance measures like the Euclidean distance and such. There is no unique definition of what a cluster is. A general intuition is that clusters should be homogeneous, often different clusters are seen as corresponding to high density areas with density gaps between different clusters.

```{r}
clusters <- hclust(dist(peo[,38:42]))
plot(clusters)
```

Hierarchical clustering (also called hierarchical cluster analysis or HCA) is a method of cluster analysis which seeks to build a hierarchy of clusters. Strategies for hierarchical clustering generally fall into two types:

Agglomerative: This is a “bottom-up” approach: each observation starts in its own cluster, and pairs of clusters are merged as one moves up the hierarchy. 

Divisive: This is a “top-down” approach: all observations start in one cluster, and splits are performed recursively as one moves down the hierarchy.

The algorithm works as follows:
1) Put each data point in its own cluster.
2) Identify the closest two clusters and combine them into one cluster.
3) Repeat the above step till all the data points are in a single cluster.

Interestingly enough the algorithm describes based on survey responses that clusters fall either into either 4 or 3 clusters. Because of the industry designation which describes that PEOs fall into 1 of 3 Tiers, we will use 3 clusters for the purpose of this test. 

```{r}
clusterCut <- cutree(clusters, 3)
table(clusterCut, peo$Tier)
```

It looks like the algorithm is not able to successfully classify the PEO tiers based on the graph above. We see that 1165 Tier 1 designations are correctly defined but the cutree method incorrectly defines cluster 4641 in other instances. I will try to hclust method to see if we can improve on these designations.

```{r}
clusters1 <- hclust(dist(peo[,38:42]), method = 'centroid')
plot(clusters1)
```

```{r}
clusterCut1 <- cutree(clusters1, 3)
table(clusterCut1, peo$Tier)
```

It looks like the centroid method improves on the complete method. We see that designations are correct more times now than incorrect for Tier 1 PEOs but still incorrectly classifies both Tier 2 and Tier 3 PEOs. We will attempt another method to determine if further improvements can be made.

```{r}
clusters2 <- hclust(dist(peo[,38:42]), method = 'ward.D')
plot(clusters2)
```


```{r}
clusterCut2 <- cutree(clusters2, 3)
table(clusterCut2, peo$Tier)
```

```{r}
clusplot(peo[,38:42],
         clusterCut1,
         lines = 0,
         shade = TRUE,
         color = TRUE,
         labels= 2,
         plotchar = FALSE,
         span = TRUE,
         main = paste('Clusters of PEOs'),
         xlab = 'Correct Rating',
         ylab = 'Incorrect Rating')
```

###Package ‘nomclust’

The nomclust() function runs hierarchical cluster analysis (HCA) with objects characterized by nominal (categorical) variables. It completely covers the clustering process, from the proximity matrix calculation to the evaluation of the quality of clustering. 

The eval component contains seven evaluation criteria in as vectors in a list. Namely, Within-cluster mutability coefficient (WCM), Within-cluster entropy coefficient (WCE), Pseudo F Indices based on the mutability (PSFM) and the entropy (PSFE), Bayessian (BIC) and Akaike (AIC) in- formation criteria for categorical data and the BK index. 

Here are the different methods being used. WCM and WCE measure the within-cluster variability in a dataset. They take values from zero to one, where values close to one indicate high variability. With increasing number of clusters, the within-cluster variability decreases, and thus, values of these indices decrease. Values of coefficients PSTau and PSU take values from zero to one as well. This time, values close to one indicate the low within-cluster variability and vice versa. Pseudo F coefficients, originally determined for quantitative data, are based on the mutability or the entropy for nominal variables.

It is important to note that coefficients based on the mutability and the entropy usually do not differ very much.
They are included in the package in order to provide two independent ways of variability computation. Big differences between mutability- and entropy-based coefficients should attract researcher’s attention.

The function contains twelve similarity measures for nominal data summarized by (Boriah et al., 2008) and by (Sulc and Rezankova, 2019). It offers three linkage methods that can be used for categorical data. The obtained clusters can be evaluated by seven evaluation criteria. The output of the nomclust() function may serve as an input for visualization functions in the nomclust package.

```{r}
# creating an object with results of hierarchical clustering of
hca.object <- nomclust(peo[,38:42], measure = "lin", method = "average",
     clu.high = 5, prox = TRUE, opt = FALSE)
```


```{r}
# obtaining values of evaluation indices
peo.eval <- hca.object$eval

# getting the optimal numbers of clusters
peo.opt <- hca.object$opt

# extracting cluster membership variables
peo.mem <- hca.object$mem

# extracting cluster membership variables as a data frame
peo.mem <- as.data.frame(hca.object$mem)
    
# obtaining a proximity matrix
peo.prox <- as.matrix(hca.object$prox)

# setting the maximal number of objects for which a proximity matrix is provided in the output to 30 
hca.object <- nomclust(peo[,38:42], measure = "lin", method = "average", clu.high = 5, prox = 30, opt = FALSE)
```

This dendrogram provides a visualization of the hierarchical relationship between the 5 clusters. We can see cluster 1 to 4 share the same parent. This is the case in real-life where PEO Tiers 1 and 2 are very closely related. It is interesting to note there is futher separation between Tier 1 PEO's.

```{r}
dend.plot(
      hca.object,
      clusters = "BIC",
      style = "greys",
      colorful = TRUE,
      clu.col = NA,
      main = "Dendrogram",
      ac = TRUE)
```

The output of the nomclust function has a form of a list with up to five components. The mem component contains cluster memberships for all cases. Because the output would be large, only first six rows of the output are displayed.

```{r}
clu_mem <- head(peo.mem) 
clu_mem
```

The output contains values for all the evaluation criteria presented in the table below for the defined range of clusters. In the first row, the WCM and WCE coefficients. The other evaluation criteria have always symbol NA in this row. The other rows represent the within-cluster variability which decreases as well. When studying the PSFM and PSFE coefficients, both of them prefer the five-cluster solution.

```{r}
clu_eval <- peo.eval
clu_eval
```

As mentioned, the eval() function provides seven evaluation criteria as a method to check on how the variability effects exist within each cluster. The criterias in this instance points to cluster 1 having the best values and diminishes from that point to the following clusters. 

```{r}
eval.plot(
  hca.object,
  criteria = "all",
  style = "greys",
  opt.col = "red",
  main = "Cluster Evaluation"
)
```


###Conclusion

In this project I used 'Nomclust' package to perform clustering analysis for determining what PEO user share similar characteristics. I introduced the main features of the nomclust package and review how closely the clusters shared characteristics with each other. This package offers a comprehensive look at the nominal clustering issue which cannot be found in any other R packages or commercial software. Hierarchical clustering of objects
characterized by nominal variables requires multiple manipulations under previous methods which exists in R. With the advent of this specific package, clustering analysis for these types of dataset can be performed efficiently. The package performed analysis well, however, requires performance optimization, because proximity matrices computation takes a lot of computational time by large datasets. In this instance, processing took approximately 4 hours to run on this dataset of 10,730 points. I have determined that the survey rankings, HealthRating, WCRating, Strategy, Cost and Support can play an important role in determining cluster formation. This in turn coincides generally with the PEO tiers.

Further analysis can be performed to gain competitive business advantage. The two areas I see where PEO providers can differentiate their services is by understanding how PEO users rank service factors within their immediate PEO competitors or to fill in niche services which corresponds to areas of under-utilization by PEO users.

####References

BORIAH, S., CHANDOLA, V. AND KUMAR, V.
Similarity Measures for Categorical Data: A Comparative Evaluation
In-text: (Boriah, Chandola and Kumar, 2008)
<http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.158.3340&rep=rep1&type=pdf> [Accessed 4 May 2021].
 
HENNIG, C.
Lecture notes for STATG019: Selected Topics in Statistics: Cluster Analysis
In-text: (Hennig, 2021)
<http://www.homepages.ucl.ac.uk/~ucakche/presentations/g19lecnotes.pdf> [Accessed 3 March 2021].

SULC, Z., CIBULKOVA, J. AND REZANKOVA, H.
Package ‘nomclust’
In-text: (Sulc, Cibulkova and Rezankova, 2019)
<https://cran.r-project.org/web/packages/nomclust/nomclust.pdf> [Accessed 5 March 2021].

SULC, Z., CIBULKOVA, J. AND REZANKOVA, H., PROCHAZJA J. 
Internal Evaluation Criteria for Categorical Data in Hierarchical Clustering: Optimal Number of Clusters Determination, Metodoloski Zveski, 15(2), p. 1-20.



